Causal Impact of Masks, Policies, Behavior on Early Covid-19 Pandemic in the U.S.

### 2020-07-16 tex/covidEffectOfNPI.tex
Version submitted in early July 2020.
Compared to submitted version the following code changes have been made:

- Removed safegraph and unacast data (they can't be publicly shared)
- Minor update to policy data loading due to data format change
- Added explicit subsetting of data to June 2, 2020 and earlier (to match what we used in submitted paper).

\begin{abstract}
This paper evaluates the dynamic impact of various policies adopted by US states on the growth rates of confirmed Covid-19  cases and deaths as well as social distancing behavior measured by Google Mobility Reports, where we take into consideration people's voluntarily behavioral response to new information of transmission risks. Our analysis finds that both policies and information on transmission risks are important determinants of Covid-19  cases and deaths and shows that a change in policies explains a large fraction of observed changes in social distancing behavior. Our counterfactual experiments suggest that nationally mandating face masks for employees on April 1st could have reduced the growth rate of cases and deaths  by more than 10 percentage points in late April, and could have led to as much as 17 to 55 percent less deaths nationally by the end of May, which roughly translates into 17 to 55 thousand saved lives.   Our estimates imply that removing non-essential business closures  (while maintaining school closures, restrictions on  movie theaters and restaurants)  could have led to -20 to 60 percent more cases and deaths by the end of May. We also find that, without stay-at-home orders, cases would have been larger by  25 to 170 percent, which implies that 0.5 to 3.4 million more Americans could have been infected if stay-at-home orders had not been implemented. Finally,  not having implemented any policies could have led to at least a 7 fold increase with an uninformative upper bound in cases (and deaths) by the end of May  in the US, with considerable uncertainty over the effects of school closures, which had little cross-sectional variation.%  (the potential range of effects could be even larger the effect of school closures on behavior is hard to identify separately from aggregate time effects).
\end{abstract}

@@

%\tableofcontents
\thispagestyle{empty}

\pagebreak

\pagestyle{empty}

@@

We present a conceptual framework that spells out the causal structure on how the Covid-19 spread  is dynamically determined by policies and human behavior. Our approach explicitly recognizes that policies not only directly affect the spread of Covid-19 (e.g., mask requirement) but also indirectly affect its spread by changing people's behavior (e.g., stay-at-home order). It also recognizes that people react to new information on Covid-19 cases and deaths, and voluntarily adjust their behavior (e.g., voluntary social distancing and hand washing) even without any policy in place.  Our casual model provides a framework to quantitatively decompose the growth of Covid-19 cases and deaths into three components: (1) direct policy effect, (2) policy effect through behavior, and (3) direct behavior effect in response to new information.\footnote{The causal model is framed using the language of structural equations models and causal diagrams of econometrics (\cite{pwright, haavelmo1944, heckman2007}; see \cite{pearl:robins}, \cite{peters2020book}, and \cite{hernanrobins2020book} for  modern developments, especially in computer science and epidemiology), with natural unfolding potential outcomes representation \citep{rubin1974,tinbergen1930,neyman:PO,imbens_rubin_2015}. As such it naturally converses in all languages for causal inference.}
% It also allows us  to explicitly define counterfactual scenarios to properly evaluate the effect of alternative policies on the spread of Covid-19.

@@

A growing number of other papers have examined the link between non-pharmaceutical interventions and Covid-19  cases.\footnote{We refer the reader to \cite{avery2020policy} for a comprehensive review of  a larger body of work researching Covid-19; here we focus on few quintessential comparisons on our work with other works that we are aware of.} \cite{hsiang2020} estimate the effect of policies on the growth rate of cases using
data from the United States, China, Iran, Italy, France, and South
Korea. In the United States, they find that the combined effect of all policies they consider on the
growth rate is $-0.347$ $(0.061)$. \cite{courtemanche2020} use US county level data to analyze the effect
of interventions on case growth rates. They find that the combination
of policies they study reduced growth rates by 9.1 percentage points
16-20 days after implementation, out of which 5.9 percentage points are attributed to shelter in place orders. Both \cite{hsiang2020} and \cite{courtemanche2020} adopted a reduced-form approach to estimate the total policy effect on case growth without using any social distancing behavior measures. In contrast, our study highlights the role of behavioral response to policies and information.

@@

Epidemiologists use model simulations to predict how cases and deaths evolve for the purpose of policy recommendation. As reviewed by \cite{avery2020policy}, there exist substantial uncertainty about the values of key epidimiological parameters \citep[see also][]{atkeson2020b,stock2020}. Simulations are often done under strong assumptions about the impact of social distancing policies without connecting to the relevant data  \citep[e.g.,][]{ferguson2020}. Furthermore, simulated models do not take into account that people may limit their contact with other people in response to higher transmission risks.\footnote{See \cite{atkeson2020a} and \cite{stock2020} for the implications of the SIR model for Covid-19 in the US. \cite{NBERw27128} estimate a SIRD model in which time-varying reproduction numbers depend on the daily deaths to capture feedback from daily deaths
to future behavior and infections.}  When such a voluntary behavioral response is ignored, simulations would produce exponential spread of disease and would over-predict  cases and deaths. Our counterfactual experiments  illustrate the importance of this voluntary behavioral change.

>>>>

\begin{abstract}
This paper evaluates the dynamic impact of various policies adopted by US states on the growth rates of confirmed Covid-19  cases and deaths as well as social distancing behavior measured by Google Mobility Reports, where we take into consideration people's voluntarily behavioral response to new information of transmission risks. Our analysis finds that both policies and information on transmission risks are important determinants of Covid-19  cases and deaths and shows that a change in policies explains a large fraction of observed changes in social distancing behavior. Our counterfactual experiments suggest that nationally mandating face masks for employees on April 1st could have reduced the growth rate of cases and deaths  by more than 10 percentage points in late April, and could have led to as much as 17 to 55 percent less deaths nationally by the end of May, which roughly translates into 17 to 55 thousand saved lives.   Our estimates imply that removing non-essential business closures  (while maintaining school closures, restrictions on  movie theaters and restaurants)  could have led to -20 to 60 percent more cases and deaths by the end of May. We also find that, without stay-at-home orders, cases would have been larger by  25 to 170 percent, which implies that 0.5 to 3.4 million more Americans could have been infected if stay-at-home orders had not been implemented. Finally,  we find considerable uncertainty over the effects of removing all policies because the timing of school closures had little cross-sectional variation.%  (the potential range of effects could be even larger the effect of school closures on behavior is hard to identify separately from aggregate time effects).
\end{abstract}

@@

%\tableofcontents

@@

We present a conceptual framework that spells out the causal structure on how the Covid-19 spread  is dynamically determined by policies and human behavior. Our approach explicitly recognizes that policies not only directly affect the spread of Covid-19 (e.g., mask requirement) but also indirectly affect its spread by changing people's behavior (e.g., stay-at-home order). It also recognizes that people react to new information on Covid-19 cases and deaths, and voluntarily adjust their behavior (e.g., voluntary social distancing and hand washing) even without any policy in place.  Our casual model provides a framework to quantitatively decompose the growth of Covid-19 cases and deaths into three components: (1) direct policy effect, (2) policy effect through behavior, and (3) direct behavior effect in response to new information.\footnote{
The causal model is framed using the language of structural equations models and causal diagrams of econometrics (\cite{pwright, haavelmo1944, heckman2007} and genetics \citep{wright1923}, with natural unfolding potential outcomes representation \citep{rubin1974,tinbergen1930,neyman:PO,imbens_rubin_2015}. See \cite{pearl:robins}, \cite{peters2020book}, and \cite{hernanrobins2020book}  for modern developments, especially in computer
science and epidemiology. The particular causal diagram has several “mediation” components, where variables affect
outcomes directly and indirectly through other variables called mediators; these structures go back at least to \cite[][see Figure 6]{wright1923}; see, e.g., \cite{baron1986} and \cite{Hines2020} for modern treatments. The father and son,
P. Wright and S. Wright closely collaborated to develop structural equation models and causal path diagrams; P.Wright’s
key work represented supply-demand system as a directed acyclical graph and established its identification using
exclusion restrictions on instrumental variables.
}
%The causal model is framed using the language of structural equations models and causal diagrams of econometrics (\cite{pwright, haavelmo1944, heckman2007}. See \cite{pearl:robins}, \cite{peters2020book}, and \cite{hernanrobins2020book} for  modern developments, especially in computer science and epidemiology), with natural unfolding potential outcomes representation \citep{rubin1974,tinbergen1930,neyman:PO,imbens_rubin_2015}. As such it naturally converses in all languages for causal inference.

% It also allows us  to explicitly define counterfactual scenarios to properly evaluate the effect of alternative policies on the spread of Covid-19.

@@

A growing number of other papers have examined the link between non-pharmaceutical interventions and Covid-19  cases.\footnote{We refer the reader to \cite{avery2020} for a comprehensive review of  a larger body of work researching Covid-19; here we focus on few quintessential comparisons on our work with other works that we are aware of.} \cite{hsiang2020} estimate the effect of policies on the growth rate of cases using
data from the United States, China, Iran, Italy, France, and South
Korea. In the United States, they find that the combined effect of all policies they consider on the
growth rate is $-0.347$ $(0.061)$. \cite{courtemanche2020} use US county level data to analyze the effect
of interventions on case growth rates. They find that the combination
of policies they study reduced growth rates by 9.1 percentage points
16-20 days after implementation, out of which 5.9 percentage points are attributed to shelter in place orders. Both \cite{hsiang2020} and \cite{courtemanche2020} adopted a reduced-form approach to estimate the total policy effect on case growth without using any social distancing behavior measures. In contrast, our study highlights the role of behavioral response to policies and information.

@@

Epidemiologists use model simulations to predict how cases and deaths evolve for the purpose of policy recommendation. As reviewed by \cite{avery2020}, there exist substantial uncertainty about the values of key epidimiological parameters \citep[see also][]{atkeson2020b,stock2020}. Simulations are often done under strong assumptions about the impact of social distancing policies without connecting to the relevant data  \citep[e.g.,][]{ferguson2020}. Furthermore, simulated models do not take into account that people may limit their contact with other people in response to higher transmission risks.\footnote{See \cite{atkeson2020a} and \cite{stock2020} for the implications of the SIR model for Covid-19 in the US. \cite{NBERw27128} estimate a SIRD model in which time-varying reproduction numbers depend on the daily deaths to capture feedback from daily deaths
to future behavior and infections.}  When such a voluntary behavioral response is ignored, simulations would produce exponential spread of disease and would over-predict  cases and deaths. Our counterfactual experiments  illustrate the importance of this voluntary behavioral change.
